---
title: "Diversity measures"
author: "George Cresswell"
date: "21/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

## Diversity measurements in the Longitudinal Metastasis Cohort

### Libraries

We use lmerTest to perform the linear mixed models. It allows us to test if there is a significant change whilst controlling for different patients.

```{r libraries, echo = TRUE, warning=FALSE, message=FALSE}
# Libraries
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())
library(lmerTest)
library(plyr)
library(reshape2)
```

### Settings for the script

```{r settings, warning=FALSE}
# Create a custom theme for plotting
plot_theme = theme_bw(base_size = 18) + theme(panel.grid.major = element_blank(), 
                                              panel.grid.minor = element_blank())

# Wide plot
wide_width = 33

# Plot dir
plot_dir = "./"
```

### Functions

Here we define functions used to measure distance. Firstly, we define `diff_state_dist` which is our measure of divergence, the number genomic bins that are different between two samples divided by total number of bins are aberrant (!=2) in at least one sample.

The second function `pga_diff_dist`, measures the difference in PGA (as percentage points).

```{r functions, warning=FALSE}
# Here we define the distance measure we will use
diff_state_dist = function(x, y, only_alt_bins = T) {
  
  frac_genome_alt = length(which(x!=y)) / length(x)
  
  if(only_alt_bins) {
    
    bins_diff = length(which(x!=y))

    c = cbind(x, y)

    bins_ab = length(which(apply(c, 1, function(i) any(i!=2))))

    output = bins_diff / bins_ab
  
  } else {output = frac_genome_alt}
  
  return(output)
  
}

# Simple difference calculation
pga_diff_dist = function(pre, post, abs = F, perc_not_frac = T) {
  
  # Calculate PGAs
  pre_pga  = length(which(pre!=2)) / length(pre)
  post_pga = length(which(post!=2)) / length(post)
  
  # Calculate increase in PGA
  output = (post_pga - pre_pga)
  
  if(perc_not_frac) {output = output * 100}
  
  if(abs) {output = abs(output)}
  
  return(output)
  
}
```

### Preprocessing

Here we process the clinical data and the CNA calls. We read in the clinical data and then the calls. We also then choose to saturate calls of losses and gain thus that 0 and 1 losses are all treated as a loss and that all gains (3, 4, 5 etc.) are treated as being simply 'gained'. This means that a change from CN 2 to 5 is treated as being identical to 2 to 1 and 3 to 1, for instance.

```{r preprocessing, warning=FALSE, message=FALSE}
# Read in clinical/genomic data
clin_genomic = read.table("Table_S7_Fig3.LM_cohort_info_v2.csv", sep = ",",
                          header = TRUE, stringsAsFactors = FALSE)

# Simplify the organs
clin_genomic$Organ[clin_genomic$Organ %in% c("Colon", "Rectum")] = "Colorectal"
clin_genomic$Organ[grep("lymph", clin_genomic$Organ)] = "Lymph"
clin_genomic$Organ[clin_genomic$Organ %in% c("Peritoneum", "Ovary", "Omentum")] = "Peritoneum"

# Split out the sample name
clin_genomic$Block  = unlist(lapply(strsplit(clin_genomic$Sample_Name, split = "-"), function(i) i[2]))
clin_genomic$Region = unlist(lapply(strsplit(clin_genomic$Sample_Name, split = "-"), function(i) i[3]))

# Read in call data
calls_list = lapply(1:nrow(clin_genomic), function(i) {
  
  # Sample info
  case   = clin_genomic[i,"Patient_ID"]
  sample = clin_genomic[i,"Sample_Name"]
  
  # Read that file in
  calls = read.table(paste0("cna_calls/",case,"_",
                            sample,"_500kb_GRCh38_cna_calls.txt"),
                     stringsAsFactors = F)
  
})

# Saturate gains and losses
calls_list = lapply(calls_list, function(m) {
  
  loss.ind = which(m[,1] <= 1)
  
  gain.ind = which(m[,1] >= 3)
  
  m[loss.ind,1] = 1
  
  m[gain.ind,1] = 3 
  
  return(m)

})

# Make them a matrix
calls_mat = do.call(cbind, calls_list)
```

```{r mode_time_points, warning=FALSE}
# Calculate number of timepoints assayed per patient
number_of_timepoint = unlist(lapply(unique(clin_genomic$Patient_ID), function(p) length(unique(clin_genomic[clin_genomic$Patient_ID==p,"Time_point"]))))

# Make a plot
hist(number_of_timepoint, xlab = "Number of timepoints assayed", main = "LM cohort sampling")
```

### Calculate divergence

We then calculate divergence for all comparisons of samples and record the organ, timepoint, type of sample (primary and metastasis) and block id for each.

```{r read_call_data, warning=FALSE}
# Run through patients
patient_dists = lapply(unique(clin_genomic$Patient_ID), function(p) {

  # Patient ids
  pat_mat = calls_mat[,which(clin_genomic$Patient_ID==p)]
  
  # Corresponding clinical information
  clin_mat = clin_genomic[which(clin_genomic$Patient_ID==p),]
  
  # Run nested apply to get all distances
  col_col_dists = lapply(1:ncol(pat_mat), function(i) {
    
    # Get column
    col_a = pat_mat[,i]
    
    # Get the organ and timepoint
    organ_i = clin_mat[i,"Organ"]
    timep_i = clin_mat[i,"Time_point"]
    type_i  = clin_mat[i,"Organ_description"]
    block_i = clin_mat[i,"Sample_Name"]
    month_i = clin_mat[i,"Month"]
    
    # Nest the other comparisons
    dists = lapply(1:ncol(pat_mat), function(j) {
      
      # Get column
      col_b = pat_mat[,j]
      
      # Get the organ and timepoint
      organ_j = clin_mat[j,"Organ"]
      timep_j = clin_mat[j,"Time_point"]
      type_j  = clin_mat[j,"Organ_description"]
      block_j = clin_mat[j,"Sample_Name"]
      month_j = clin_mat[j,"Month"]
      
      # Get distance
      d = diff_state_dist(x = col_a, y = col_b)
      p = pga_diff_dist(pre = col_a, post = col_b, abs = T, perc_not_frac = F)
      
      # Give NA if it is the same comparison or redundant
      if(i<=j) {d = NA}
      if(i<=j) {p = NA}
      
      # Make the output a dataframe
      output = data.frame(dist = d,
                          Organ_i = organ_i,
                          Organ_j = organ_j,
                          TP_i = timep_i,
                          TP_j = timep_j,
                          Type_i = type_i,
                          Type_j = type_j,
                          Code_i = block_i,
                          Code_j = block_j,
                          Month_i = month_i,
                          Month_j = month_j,
                          pdiff = p)
      
      return(output)
      
    })
    
    # Get those dists
    return(dists)
    
  })
  
  # Dist mat
  one_side_dist_mat = do.call(rbind, lapply(col_col_dists, function(i) do.call(rbind, i)))
  
  # One sided dist_mat for patient
  return(one_side_dist_mat)
  
})

# Name elements in the list
names(patient_dists) = unique(clin_genomic$Patient_ID)

# Also add it as a column
for(i in 1:length(patient_dists)) {
  
  # Add list name as a column
  patient_dists[[i]]$Patient = names(patient_dists)[i]
  
}

# Remove NAs
patient_dists = lapply(patient_dists, na.omit)

# Make into one massive df
cohort_dist_m = do.call(rbind, patient_dists)
```

### Mean single timepoint divergence

Next, we calculate the *mean* divergence in each single time interval for each patient and we report the median value. This is provide a value for the general amount of divergence measured.

```{r mean_divergence_cohort_tps, warning=FALSE}
# Calculate average per patient and timepoint difference
per_patient_tp_divergence = lapply(unique(cohort_dist_m$Patient), function(p) {
  
  # Take patient
  pat_m =  cohort_dist_m[cohort_dist_m$Patient==p,]
  
  #Subset for single difference timepoints
  pat_m = pat_m[abs(pat_m$TP_i - pat_m$TP_j) == 1,]
  
  # Make sure the timepoints are in order
  TPs = apply(pat_m, 1, function(r) paste0(sort(c(r["TP_i"],r["TP_j"])), collapse = "_"))
  
  # What are these unique TPs?
  diff_tps = unique(TPs)
  
  # Run through the tps
  per_tp = lapply(diff_tps, function(t) {
    
    # Calculate mean
    mean_div_cohort = mean(pat_m[TPs==t,"dist"])
    
  })
  
  unlist(per_tp)
  
})

# Print some infomation to the user
print(paste0("Median mean timepoint divergence: ",round(median(unlist(per_patient_tp_divergence)), digits = 3)))
```

### Within timepoint, same organ divergence

Here we plot the divergences observed within each single timepoint, within a single organ, to represent divergence in space.

```{r divergences_in_timepoint_in_organs, warning=FALSE, fig.width=33, fig.height=6}
# Remove anything that isn't timepoint to timepoint comparison
plot.df = cohort_dist_m[cohort_dist_m$Organ_i==cohort_dist_m$Organ_j & cohort_dist_m$TP_i==cohort_dist_m$TP_j,]

# Remove ugly rownames
rownames(plot.df) = NULL

# Make factor
plot.df$Timepoint = as.factor(plot.df$TP_i)

# Extract the month
plot.df$Month = as.factor(plot.df$Month_i)

# Do ggplot2
ggplot(plot.df, aes(x = Month, y = dist)) + geom_violin() + 
  ylab("Divergence (fraction bins altered)") + xlab("Month") +
  geom_jitter(width = 0.1, aes(col = Organ_i)) + ylim(-0.1,1) + #geom_hline(yintercept = 0.2) +
  facet_grid(. ~ Patient, scales = "free_x") + plot_theme + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(col = "Organ")
ggsave(filename = paste0(plot_dir,
                         "Divergences_within_timepoints_within_organs.pdf"), 
       width = wide_width, height = 6)

# Same plot but for the PGA difference
ggplot(plot.df, aes(x = Month, y = pdiff)) + geom_violin() + 
  ylab("PGA difference") + xlab("Month") +
  geom_jitter(width = 0.1, aes(col = Organ_i)) + ylim(-0.1,1) + #geom_hline(yintercept = 0.2) +
  facet_grid(. ~ Patient, scales = "free_x") + plot_theme + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(col = "Organ")
```

### Same timepoint, difference organ divergence

Here we plot the divergences observed within each single timepoint, but this time within different organs sampled at these timepoints, and this represents divergence in space between the samples across the body. There are very few examples of timepoints in which multiple organs were assessed. Generally in the cohort colorectal samples are sampled in the first timepoint and liver in latter timepoints.

```{r between_organs_within_timepoints, warning=FALSE, fig.width=15, fig.height=6}
# Plotting divergence in a timepoint between organs
plot.df = do.call(rbind, lapply(patient_dists, function(p_data) {
  
  p_data[which(p_data$TP_i==p_data$TP_j & p_data$Organ_i!=p_data$Organ_j),]
  
}))

# Remove ugly rownames
rownames(plot.df) = NULL

# Make factor
plot.df$Timepoint = as.factor(plot.df$TP_i)

# We need it as a factor
plot.df$Month = as.factor(plot.df$Month_i)

# Add on, for colours
plot.df$Organ_combination = as.factor(apply(t(apply(plot.df, 1, function(r) sort(c(r[2], r[3])))), 1, 
                                            function(i) paste0(i, collapse = "|")))

# Do ggplot2
ggplot(plot.df, aes(x = Month, y = dist)) + geom_violin() + 
  ylab("Divergence (fraction bins altered)") + xlab("Month") +
  geom_jitter(width = 0.1, aes(col = Organ_combination)) + ylim(-0.1,1) + 
  facet_grid(. ~ Patient, scales = "free_x") + plot_theme + labs(col = "Organ Comparison")
ggsave(filename = paste0(plot_dir,
                         "Divergences_between_organs_within_timepoints.pdf"), 
       width = 15, height = 6)

# Same thing but for PGA difference
ggplot(plot.df, aes(x = Month, y = pdiff)) + geom_violin() + 
  ylab("PGA difference") + xlab("Month") +
  geom_jitter(width = 0.1, aes(col = Organ_combination)) + ylim(-0.1,1) + 
  facet_grid(. ~ Patient, scales = "free_x") + plot_theme + labs(col = "Organ Comparison")
```

### Within organ divergence, primary and metastasis

Now we collapse this data and group the within timepoint, single organ divergences into primary and metastasis categories to investigate if metastatic deposits show a higher or lower level of within organ diversity than primary tumours. Here we can also see that we have much more data for liver metastases. 

```{r make_df_within_prim_mets, warning=FALSE, fig.width = 8, fig.height = 6}
# Get only the heterogeneities within primaries and mets
within_prim_mets = cohort_dist_m[cohort_dist_m$TP_i==cohort_dist_m$TP_j & 
                                 cohort_dist_m$Type_i==cohort_dist_m$Type_j & 
                                 cohort_dist_m$Organ_i==cohort_dist_m$Organ_j,]

# Reorder those levels
within_prim_mets$Type_i = factor(within_prim_mets$Type_i, levels(within_prim_mets$Type_i)[2:1])
```

### Mixed model of within organ divergence, primary and metastasis

Of course, to correctly investigate this we need to perform some statistics. Therefore here we create a mixed model (`lmer`) of `Divergence ~ Metastasis` where we use each patient as a random effect to control for the fact that patients may show different total levels of divergence.

```{r mixed_model_primary_met, echo = FALSE, warning=FALSE}
# Do mixed modelling
res2b = lmer(dist ~ Type_i + (1 | Patient), 
             REML = TRUE, data = within_prim_mets)
print("Mixed model (1) - Primary and Metastasis using patients as random intercepts")
summary(res2b)

# P Values
p_value_prim_met_diff = round(summary(res2b)$coefficients["Type_iMetastasis","Pr(>|t|)"], digits = 2)
p_value_line_height = max(within_prim_mets$dist)*1.1

# Plot divergences in Mets and Primary
ggplot(within_prim_mets, aes(x = Type_i, y = dist)) + geom_boxplot() +
  ylab("Divergence (fraction bins altered)") + xlab(NULL) +
  geom_jitter(width = 0.2, aes(col = Patient, shape = Organ_i), size = 3) + ylim(-0.1,1) + plot_theme + 
  theme(text = element_text(size = 20)) + 
  geom_segment(aes(x = 1, y = p_value_line_height, xend = 2, yend = p_value_line_height)) +
  geom_segment(aes(x = 1, y = p_value_line_height - 0.01, xend = 1, yend = p_value_line_height)) +
  geom_segment(aes(x = 2, y = p_value_line_height - 0.01, xend = 2, yend = p_value_line_height)) +
  annotate("text", label = paste0("p=",p_value_prim_met_diff), x = 1.5, y = p_value_line_height + 0.075, size = 5) +
  scale_shape_manual(values=c(17, 19, 15, 18, 4)) +
  labs(col = "Patient", shape = "Organ (Panel C)")
ggsave(filename = paste0(plot_dir,
                         "Divergences_within_organs_categorised_by_primary_metastatsis.png"), 
       width = 7, height = 8.2)

rownames(within_prim_mets) = NULL

# Print mean divergence in the primary samples
print(mean(within_prim_mets[within_prim_mets$Type_i=="Primary","dist"]))

# Do mixed modelling
res2b = lmer(pdiff ~ Type_i + (1 | Patient), 
             REML = TRUE, data = within_prim_mets)
print("Repeat of mixed model (1) but for PGA difference")
summary(res2b)

# P Values
p_value_prim_met_diff = round(summary(res2b)$coefficients["Type_iMetastasis","Pr(>|t|)"], digits = 2)
p_value_line_height = max(within_prim_mets$dist)*1.1

# Plot PGA differences in Mets and Primary
ggplot(within_prim_mets, aes(x = Type_i, y = pdiff)) + geom_violin() +
  ylab("PGA difference") + xlab("") +
  geom_jitter(width = 0.1, aes(col = Organ_i)) + ylim(-0.1,1) + plot_theme + theme(text = element_text(size = 24)) + 
  geom_segment(aes(x = 1, y = p_value_line_height, xend = 2, yend = p_value_line_height)) +
  geom_segment(aes(x = 1, y = p_value_line_height - 0.01, xend = 1, yend = p_value_line_height)) +
  geom_segment(aes(x = 2, y = p_value_line_height - 0.01, xend = 2, yend = p_value_line_height)) +
  annotate("text", label = paste0("p=",p_value_prim_met_diff), x = 1.5, y = p_value_line_height + 0.075, size = 5) +
  labs(col = "Organ")
```

### Compare divergence within colorectal and liver samples and between them

Now we would like to ask if there is a significant difference between divergence within liver and colorectal samples from the same patient and between them. For this we do not filter by timepoint. As we can see there are far fewer colorectal-colorectal comparisons compared to anything to the liver samples.

### ANOVA for difference between intra- and inter- organ variation

So now we ask if any of the categories have a significantly different mean using a simple one-way ANOVA (`aov`).

```{r compare_div_colorectal_liver, warning=FALSE, fig.width=12, fig.height=6}
# Collapse the organ comparisons
organ_comparison = apply(cohort_dist_m, 1, function(r) {
  
  i = as.character(r[2])
  j = as.character(r[3])
  
  c = c(i,j)
  
  x = paste(c[order(c)], collapse = "|")

})

# Make a unique comparison
comparison_organs = data.frame(cohort_dist_m, Organ_Compare = organ_comparison)

# Subset only colorectal and liver
comparison_organs = comparison_organs[comparison_organs$Organ_Compare %in% 
                                      c("Colorectal|Colorectal", "Liver|Liver", 
                                        "Colorectal|Liver"),]
comparison_organs$Organ_Compare = droplevels(comparison_organs$Organ_Compare)

# Here I will do an ANOVA first for divergence
print("An ANOVA for comparing divergence in colorectal and liver organs...")
anova_div = aov(dist ~ Organ_Compare, comparison_organs)
summary(anova_div)
tukey_divergence = TukeyHSD(anova_div)

# Make a look up table to plot p-values
p_value_coord = as.data.frame(do.call(rbind, strsplit(rownames(tukey_divergence$Organ_Compare), "-")), 
                              stringsAsFactors = F)

# Add in adjusted p values
p_value_coord$p_val_adj = tukey_divergence$Organ_Compare[,"p adj"]

# Now I will do an ANOVA first for PGA difference
print("An ANOVA for comparing difference in PGA in colorectal and liver organs...")
summary(aov(pdiff ~ Organ_Compare, comparison_organs))

# Make a ggplot
ggplot(comparison_organs, aes(x = Organ_Compare, y = dist)) + geom_boxplot() + 
  ylab("Divergence (fraction bins altered)") + xlab("") +
  geom_jitter(width = 0.1, aes(col = Patient)) + ylim(-0.1,1) + 
  geom_segment(aes(x = p_value_coord$V1[1], y = 0.85, 
                   xend = p_value_coord$V2[1], yend = 0.85)) +
  annotate("text", label = paste0("p=",round(p_value_coord$p_val_adj[1], digits = 2)), x = 1.5, y = 0.85 + 0.03, size = 4) +
  geom_segment(aes(x = p_value_coord$V1[2], y = 0.9, 
                   xend = p_value_coord$V2[2], yend = 0.9)) +
  annotate("text", label = paste0("p=",round(p_value_coord$p_val_adj[2], digits = 2)), x = 2, y = 0.9 + 0.03, size = 4) +
  geom_segment(aes(x = p_value_coord$V1[3], y = 0.95, 
                   xend = p_value_coord$V2[3], yend = 0.95)) +
  annotate("text", label = paste0("p=",round(p_value_coord$p_val_adj[3], digits = 2)), x = 2.5, y = 0.95 + 0.03, size = 4) +
  annotate("text", label = paste0("Tukey HSD test"), x = 2, y = 1, size = 4) +
  plot_theme + theme(text = element_text(size = 24)) + labs(col = "Patient")
ggsave(filename = paste0(plot_dir,
                         "Non_timepoint_specific_colorectal_liver_comparison.pdf"), 
       width = 12, height = 6)

# Repeat for PGA difference
ggplot(comparison_organs, aes(x = Organ_Compare, y = pdiff)) + geom_violin() + 
  ylab("PGA difference") + xlab("") +
  geom_jitter(width = 0.1, aes(col = Patient)) + ylim(-0.1,1) + plot_theme + 
  theme(text = element_text(size = 24)) + labs(col = "Patient")
```

### Divergence found within same timepoint and in different timepoints

Now we want to know if divergence comparisons within timepoints are different from those not in the same timepoint. We now plot this simple category.

To assess this we simply ask what is the effect on divergence if a comparison is made within the same timepoint or not. We again do a mixed model (`lmer`) using the patient as a random effect.

```{r within_tp_across_tps, warning=FALSE, fig.width=10, fig.height=6}
# Now we look into comparing within timepoint and across timepoint samples
within_tps_across_tps = data.frame(cohort_dist_m, Same_tp = cohort_dist_m$TP_i==cohort_dist_m$TP_j)

# Do mixed modelling
res2b = lmer(dist ~ Same_tp + (1 | Patient), 
             REML = TRUE, data = within_tps_across_tps)
print("Mixed model (2) - General assessment of whether within timepoint and across timepoint is different")
summary(res2b)

# Still for plotting a pvalue
p_value_same_tp = signif(summary(res2b)$coefficients["Same_tpTRUE","Pr(>|t|)"], digits = 2)
p_value_line_height = max(within_tps_across_tps$dist)*1.1

# Make a ggplot
ggplot(within_tps_across_tps, aes(x = Same_tp, y = dist)) + geom_violin() + 
  ylab("Divergence (fraction bins altered)") + xlab("Same timepoint?") +
  geom_jitter(width = 0.1, aes(col = Patient)) + ylim(-0.1,1) + plot_theme + 
  geom_segment(aes(x = 1, y = p_value_line_height, xend = 2, yend = p_value_line_height)) +
  geom_segment(aes(x = 1, y = p_value_line_height - 0.01, xend = 1, yend = p_value_line_height)) +
  geom_segment(aes(x = 2, y = p_value_line_height - 0.01, xend = 2, yend = p_value_line_height)) +
  annotate("text", label = paste0("p=",p_value_same_tp), x = 1.5, y = p_value_line_height + 0.075, size = 5) +
  theme(text = element_text(size = 24)) + labs(col = "Patient")
ggsave(filename = paste0(plot_dir,
                         "General_same_timepoint_not_same_comparison.pdf"), 
       width = 10, height = 6)

# What does a mixed linear model look like for PGA difference
res2b = lmer(pdiff ~ Same_tp + (1 | Patient),
             REML = TRUE, data = within_tps_across_tps)
print("Repeat of mixed model (2) but for PGA difference")
summary(res2b)

# Still for plotting a pvalue
p_value_same_tp = signif(summary(res2b)$coefficients["Same_tpTRUE","Pr(>|t|)"], digits = 2)
p_value_line_height = max(within_tps_across_tps$dist)*1.1

# Repeat for PGA difference
ggplot(within_tps_across_tps, aes(x = Same_tp, y = pdiff)) + geom_violin() + 
  ylab("PGA Difference") + xlab("Same timepoint?") +
  geom_jitter(width = 0.1, aes(col = Patient)) + ylim(-0.1,1) + plot_theme + 
  geom_segment(aes(x = 1, y = p_value_line_height, xend = 2, yend = p_value_line_height)) +
  geom_segment(aes(x = 1, y = p_value_line_height - 0.01, xend = 1, yend = p_value_line_height)) +
  geom_segment(aes(x = 2, y = p_value_line_height - 0.01, xend = 2, yend = p_value_line_height)) +
  annotate("text", label = paste0("p=",p_value_same_tp), x = 1.5, y = p_value_line_height + 0.075, size = 5) +
  theme(text = element_text(size = 24)) + labs(col = "Patient")
```

### Divergences between timepoint, including time normalised

Now we are interested in those divergences between single intervals across the patients both as a total value and also normalised by time (divergence month^-1^).

```{r time_normalised_analysis, warning=FALSE, fig.width=33, fig.height=6}
# Calculate date difference
patient_dists = lapply(patient_dists, function(p) {
  
  months_annot = lapply(1:nrow(p), function(r) {
    
    m1 = p[r,"Month_i"]
    
    m2 = p[r,"Month_j"]
    
    ms = sort(c(m1, m2))
    
    month_annot = paste0("Month_",ms[1],"_to_",ms[2])
    
    return(month_annot)
    
  })
  
  steps_away = lapply(1:nrow(p), function(r) {
    
    tp1     = p[r,"TP_i"]
    
    tp2     = p[r,"TP_j"]
    
    steps = abs(tp2 - tp1)
    
    return(steps)
    
  })
  
  treatment_prior_to_most_recent_tp = lapply(1:nrow(p), function(r) {
    
    patient = p[r,"Patient"]
    
    tp1     = p[r,"TP_i"]
    
    tp2     = p[r,"TP_j"]
    
    max_tp  = max(tp1, tp2)
    
    prev_tr = unique(clin_genomic[which(clin_genomic$Patient_ID==patient & 
                                        clin_genomic$Time_point==max_tp),"Treatment_in_previous_interval"])
    
    return(prev_tr)
    
  })
  
  p$Difference_points = unlist(steps_away)
  
  p$Treatment_prior_recent_tp = unlist(treatment_prior_to_most_recent_tp)
  
  p$Month_annotation = unlist(months_annot)
  
  return(p)
  
})

# Collapse the full data set
patient_dists_mat = do.call(rbind, patient_dists)

# Remove ugly row names
rownames(patient_dists_mat) = NULL

# Write it out
write.table(patient_dists_mat, 
            file = paste0(plot_dir,"LM_cohort_divergence_comparison.txt"), 
            sep = ",", row.names = F, quote = F)

# Remove anything that is not a comparison of one timepoint to the next
patient_dists_next_tp = lapply(patient_dists, function(p) p[p$Difference_points==1,])

# Collapse and normal
patient_dists_next_tp = do.call(rbind, patient_dists_next_tp)

# Time comparison
patient_dists_next_tp$Timepoint_comparison = apply(patient_dists_next_tp, 1, 
                                                   function(r) paste0(sort(r[4:5]), collapse = "_"))

# Add space instead of underscore
patient_dists_next_tp$Month_annotation = gsub("_", " ", patient_dists_next_tp$Month_annotation)

# Make month annotation a factor
patient_dists_next_tp$Month_annotation = as.factor(patient_dists_next_tp$Month_annotation)

# The order first month
level_order = order(as.numeric(unlist(lapply(strsplit(levels(patient_dists_next_tp$Month_annotation), 
                                                      split = " "), function(i) i[2]))))

# Reorder them levels
patient_dists_next_tp$Month_annotation = factor(patient_dists_next_tp$Month_annotation, 
                                                levels(patient_dists_next_tp$Month_annotation)[level_order])

# Do ggplot2
ggplot(patient_dists_next_tp, aes(x = Month_annotation, y = dist)) + geom_violin() + 
  ylab("Divergence (fraction bins altered)") + xlab("Timepoint comparison") +
  geom_jitter(width = 0.1, aes(col = Treatment_prior_recent_tp)) + ylim(-0.1,1) + 
  facet_grid(. ~ Patient, scales = "free_x", space = "free_x") + plot_theme + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(col = "Treatment in interval")
ggsave(filename = paste0(plot_dir,"Divergences_between_timepoints_total.pdf"), 
       width = wide_width, height = 6)

# Add one for PGA difference
ggplot(patient_dists_next_tp, aes(x = Month_annotation, y = pdiff)) + geom_violin() + 
  ylab("PGA difference") + xlab("Timepoint comparison") +
  geom_jitter(width = 0.1, aes(col = Treatment_prior_recent_tp)) + ylim(-0.1,1) + 
  facet_grid(. ~ Patient, scales = "free_x", space = "free_x") + plot_theme + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(col = "Treatment in interval")
```

### Inter-timepoint divergence categorised by treatment status

Now we are interested in whether treatment affects these intervals, so we basically collapse the previous plot and plot divergence both normalised and not normalised by time in the off- and on- treatment categories.

### Mixed model of non-time normalised divergence on and off treatment

Here we perform a mixed model to test if absolute differences in divergence are different between treatments, using patients as a random effect.

```{r mixed_model_divergence_treated, warning=FALSE, fig.width=10, fig.height=7}
# Give informative name
patient_dists_next_tp$Treatment_name = c("No Treatment", 
                                         "Treated")[patient_dists_next_tp$Treatment_prior_recent_tp + 1]

# Do linear modelling
res2b = lmer(dist ~ Treatment_prior_recent_tp + (1 | Patient), 
             REML = TRUE, data = patient_dists_next_tp)
print("Mixed model (3) - Divergence NOT normalised by time in treated and non-treated intervals using patients as random intercepts")
summary(res2b)

# Still for plotting a pvalue
p_value_treatment_diff = round(summary(res2b)$coefficients["Treatment_prior_recent_tpTRUE","Pr(>|t|)"], digits = 2)
p_value_line_height = max(patient_dists_next_tp$dist)*1.1

# Do ggplot2
ggplot(patient_dists_next_tp, aes(x = Treatment_name, y = dist)) + geom_boxplot() + 
  ylab("Divergence (fraction bins altered)") +
  geom_jitter(width = 0.2, aes(col = Patient), size = 3) + ylim(-0.1,1) + xlab(NULL) +
  geom_segment(aes(x = 1, y = p_value_line_height, xend = 2, yend = p_value_line_height)) +
  geom_segment(aes(x = 1, y = p_value_line_height - 0.01, xend = 1, yend = p_value_line_height)) +
  geom_segment(aes(x = 2, y = p_value_line_height - 0.01, xend = 2, yend = p_value_line_height)) +
  annotate("text", label = paste0("p=",p_value_treatment_diff), x = 1.5, y = p_value_line_height + 0.075, size = 5) +
  plot_theme + theme(text = element_text(size = 20)) + labs(col = "Patient")
ggsave(filename = paste0(plot_dir,
                         "Divergences_between_timepoints_total_chemo_status.png"), 
       width = 7, height = 8.2)

# Also take the opportunity to test it with a linear mixed model
res2b = lmer(pdiff ~ Treatment_prior_recent_tp + (1 | Patient), 
             REML = TRUE, data = patient_dists_next_tp)
print("Repeat of mixed model (3) but for PGA difference")
summary(res2b)

# Still for plotting a pvalue
p_value_treatment_diff = round(summary(res2b)$coefficients["Treatment_prior_recent_tpTRUE","Pr(>|t|)"], digits = 2)
p_value_line_height = max(patient_dists_next_tp$dist)*1.1

# Again let's investigate PGA difference in the same comparison
ggplot(patient_dists_next_tp, aes(x = Treatment_name, y = pdiff)) + geom_violin() + 
  ylab("PGA difference") + xlab("Treatment in interval") +
  geom_jitter(width = 0.1, aes(col = Patient)) + ylim(-0.1,1) + xlab("") +
  geom_segment(aes(x = 1, y = p_value_line_height, xend = 2, yend = p_value_line_height)) +
  geom_segment(aes(x = 1, y = p_value_line_height - 0.01, xend = 1, yend = p_value_line_height)) +
  geom_segment(aes(x = 2, y = p_value_line_height - 0.01, xend = 2, yend = p_value_line_height)) +
  annotate("text", label = paste0("p=",p_value_treatment_diff), x = 1.5, y = p_value_line_height + 0.075, size = 5) +
  plot_theme + theme(text = element_text(size = 24)) + labs(col = "Patient")

```

### Average change in PGA across timepoints

Here we process the data to calculate percentage of the genome aberrated (PGA) and to compare from one timepoint to the next whether this goes up or down and record the difference. Here is PGA in one sample is 0.41 and in the next it is 0.43, we would report it as going up by 0.02 * 100 (2). We average this for each timepoint and report the median of the averages across the cohort.

```{r pga_change_timepoints, warning=FALSE}
######################################
### Task: Calculate changes in PGA ###
######################################

# Run through patients
patient_pga_change = lapply(unique(clin_genomic$Patient_ID), function(p) {
  
  # Patient ids
  pat_mat = calls_mat[,which(clin_genomic$Patient_ID==p)]
  
  # Corresponding clinical information
  clin_mat = clin_genomic[which(clin_genomic$Patient_ID==p),]
  
  # Run nested apply to get all distances
  col_col_dists = lapply(1:ncol(pat_mat), function(i) {
    
    # Get column
    col_a = pat_mat[,i]
    
    # Get the organ and timepoint
    organ_i = clin_mat[i,"Organ"]
    timep_i = clin_mat[i,"Time_point"]
    type_i  = clin_mat[i,"Organ_description"]
    block_i = clin_mat[i,"Sample_Name"]
    
    # Nest the other comparisons
    dists = lapply(1:ncol(pat_mat), function(j) {
      
      # Get column
      col_b = pat_mat[,j]
      
      # Get the organ and timepoint
      organ_j = clin_mat[j,"Organ"]
      timep_j = clin_mat[j,"Time_point"]
      type_j  = clin_mat[j,"Organ_description"]
      block_j = clin_mat[j,"Sample_Name"]
      
      # Get distance
      d = pga_diff_dist(pre = col_a, post = col_b, abs = T, perc_not_frac = T)
      
      # Give NA if it is the same comparison
      if(timep_i!=(timep_j-1)) {d = NA}
      
      # Take the entry of the two which has the most recent timepoint and ask if it was treated in the interval
      #   Non-sequential comparisons will be removed by na.omit due to above statement
      treated = clin_mat[c(i, j)[which.max(c(timep_i, timep_j))],"Treatment_in_previous_interval"]
      
      # Make the output a dataframe
      output = data.frame(pdiff = d,
                          Organ_i = organ_i,
                          Organ_j = organ_j,
                          TP_i = timep_i,
                          TP_j = timep_j,
                          Type_i = type_i,
                          Type_j = type_j,
                          Code_i = block_i,
                          Code_j = block_j,
                          Treated = treated)
      
      return(output)
      
    })
    
    # Get those dists
    return(dists)
    
  })
  
  # Dist mat
  one_side_dist_mat = do.call(rbind, lapply(col_col_dists, function(i) do.call(rbind, i)))
  
  # One sided dist_mat for patient
  return(one_side_dist_mat)
  
})

# Name elements in the list
names(patient_pga_change) = unique(clin_genomic$Patient_ID)

# Also add it as a column
for(i in 1:length(patient_pga_change)) {
  
  # Add list name as a column
  patient_pga_change[[i]]$Patient = names(patient_pga_change)[i]
  
}

# Remove NAs
patient_pga_change = lapply(patient_pga_change, na.omit)

# Make into one massive df
pga_change_m = do.call(rbind, patient_pga_change)

# Calculate average per patient and timepoint difference
per_patient_tp_pga_change = lapply(unique(pga_change_m$Patient), function(p) {
  
  # Take patient
  pat_m =  pga_change_m[pga_change_m$Patient==p,]
  
  # What timepoints are there in this patient?
  diff_tps = unique(paste0(pat_m[,"TP_i"],"_",pat_m[,"TP_j"]))
  
  # Run through the tps
  per_tp = lapply(diff_tps, function(t) {
    
    # Calculate mean
    mean_pga_change = mean(pat_m[paste0(pat_m[,"TP_i"],"_",pat_m[,"TP_j"])==t,"pdiff"])
    
    # Was this interval treated?
    treated = unique(pat_m[paste0(pat_m[,"TP_i"],"_",pat_m[,"TP_j"])==t,"Treated"])
    
    # Catch for error 
    if(length(treated)!=1) {stop("Multiple treatment in interval statements for same timepoint")}
    
    # Return treatment status and mean pga change
    out = data.frame(Mean_PGA_change = mean_pga_change, 
                     Treated = treated, Interval =t, 
                     Patient = p)
    
    return(out)
    
  })
  
  do.call(rbind, per_tp)
  
})

# Collapse into a full table
per_patient_tp_pga_change = do.call(rbind, per_patient_tp_pga_change)

# What is the median change in PGA in all time intervals
print(paste0("Median mean increase in PGA (+ / -): ",
             round(median(per_patient_tp_pga_change$Mean_PGA_change), digits = 1)))

# Variable ylim max
ylim_max = round(max(per_patient_tp_pga_change$Mean_PGA_change) + 10, digits = -1)

# Make a histogram
ggplot(per_patient_tp_pga_change, aes(x = Treated, y = Mean_PGA_change)) +
  geom_hline(yintercept = median(per_patient_tp_pga_change$Mean_PGA_change)) +
  geom_boxplot(width=0.5) + ylab("Mean PGA change in interval") + ylim(0,ylim_max) + 
  xlab("Treated in interval") + 
  ggtitle("Mean divergence between all sequential timepoints", subtitle = "(horizontal line displays whole cohort median)") + 
  plot_theme + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
ggsave(filename = paste0(plot_dir,
                         "Mean_PGA_change_timepoints_treatment.png"), 
       width = 9, height = 7)
  
```

### Mean PGA across time

```{r timeline_pga, warning=FALSE}
# Make new object
clin_genomic2 = clin_genomic

# Melt to calculate means of PGAs
melted_pga_means = ddply(melt(clin_genomic2[,c("Patient_ID", "Time_point", 
                                               "Days", "Treatment_in_previous_interval",
                                               "PGA")], 
                         id.vars = c("Patient_ID", "Time_point", "Days", 
                                     "Treatment_in_previous_interval")), 
                         .(Patient_ID, Time_point, Days, 
                           Treatment_in_previous_interval, variable), summarize, mean = mean(value))

# Add info on whether it is going up or down
melted_pga_means$change_in_next = unlist(lapply(1:nrow(melted_pga_means), function(r) {
    
    p = melted_pga_means[r,"Patient_ID"]
    
    t = melted_pga_means[r,"Time_point"]
    
    d = melted_pga_means[which(melted_pga_means$Patient_ID==p &  melted_pga_means$Time_point==(t+1)),"mean"] - melted_pga_means[r,"mean"]
    
    if(length(d)==0) {d = NA}
    
    return(d)
    
}))

# What threshold of increase to use?
change_threshold = 0.1

# Categorise and significantly up or down
melted_pga_means$change_in_pga[melted_pga_means$change_in_next >=  change_threshold] = "Up"
melted_pga_means$change_in_pga[melted_pga_means$change_in_next <= -change_threshold] = "Down"

# Print percentage points change
print(paste0("Percentage points change: ",round(mean(melted_pga_means$change_in_next, na.rm = T)*100, digits = 2)))

ggplot(melted_pga_means, aes(x = Days+1, y = mean*100)) + geom_point() + ylab("Mean PGA at timepoint") +
  xlab("Days since first timepoint") +
  geom_line(aes(group = Patient_ID)) + ylim(0,100)

ggplot(melted_pga_means, aes(x = Days+1, y = mean*100)) + geom_point(size = 3) + ylab("Mean PGA at timepoint") +
  xlab("Day of time course") + scale_x_continuous(trans='log10', breaks = c(1, 10, 100, 1000)) +
  geom_line(aes(group = Patient_ID)) + ylim(0,100) + theme(text = element_text(size = 20), legend.position = "none")
ggsave(filename = paste0(plot_dir,
                         "Mean_PGA_timepoints_patients.png"), 
       width = 7, height = 8.2)
```
